What We Eval in the Shadows

### 2021-04-12 analysis/corpus.Rmd

eval_plot_df %>% 
#  filter(evals > 0, evals_hit > 0) %>%
  mutate(
    # size=cut(package_r_code, size_breaks, size_labels, include.lowest=T),
    label=ifelse(evals %in% tail(sort(evals),3), package, NA)
  ) %>%
  ggplot(aes(x=evals, y=evals_hit, size=package_r_code, label=label)) + 
    geom_jitter(alpha=0.3) +
    geom_text(nudge_y = .25, check_overlap = TRUE, size=4) +
    scale_x_continuous(
      trans="log10",
      breaks=function(x) {
        f <- scales::breaks_log()
        sort(c(f(x), max(eval_plot_df$evals)))
      },
      labels=scales::comma_format(accuracy = 1)
    ) +
    scale_y_continuous(
      trans="pseudo_log",
      breaks=function(x) {
        f <- scales::breaks_log()
        sort(c(0, 1, f(c(1, x[2])), max(eval_plot_df$evals_hit)))
      },
      labels=scales::comma_format(accuracy = 1)
    ) +
    scale_size_binned(
      range=c(1,10),
      breaks = size_breaks,
      labels=scales::comma_format(accuracy = 1)
    ) +
#    geom_smooth(method = "lm", se = FALSE, show.legend = FALSE, linetype="dashed", color="red", size=0.2) +
    labs(x="Eval call sites (log)", y="Traced eval call sites (log)", size="Package R code size") +
    theme(
      text = element_text(size=17),
      legend.position = c(0.15, 0.8)
    )
ggsave(path(PLOT_DIR, "traced-eval-callsites.pdf"))
```

>>>>

eval_plot_df %>% 
#  filter(evals > 0, evals_hit > 0) %>%
  mutate(
    # size=cut(package_r_code, size_breaks, size_labels, include.lowest=T),
    label=ifelse(evals %in% tail(sort(evals),3), package, NA)
  ) %>%
  ggplot(aes(x=evals, y=evals_hit, size=package_r_code, label=label)) + 
    geom_jitter(alpha=0.3) +
    geom_text(nudge_y = .25, check_overlap = TRUE, size=4) +
    scale_x_continuous(
      trans="log10",
      breaks=function(x) {
        f <- scales::breaks_log()
        sort(c(f(x), max(eval_plot_df$evals)))
      },
      labels=scales::comma_format(accuracy = 1)
    ) +
    scale_y_continuous(
      trans="pseudo_log",
      breaks=function(x) {
        f <- scales::breaks_log()
        sort(c(0, 1, f(c(1, x[2])), max(eval_plot_df$evals_hit)))
      },
      labels=scales::comma_format(accuracy = 1)
    ) +
    scale_size_binned(
      range=c(1,10),
      breaks = size_breaks,
      labels=scales::comma_format(accuracy = 1)
    ) +
#    geom_smooth(method = "lm", se = FALSE, show.legend = FALSE, linetype="dashed", color="red", size=0.2) +
    labs(x="Eval call sites (log)", y="Traced eval call sites (log)", size="Package R code size") +
    theme(
      text = element_text(size=17),
      legend.position = c(0.2, 0.75)
    )
ggsave(path(PLOT_DIR, "traced-eval-callsites.pdf"))
```

### 2021-04-12 analysis/side-effects.Rmd

---
title: "Side effects"
output: html_document
params:
  base_dir: /home/aviral/projects/revalstudy2
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
for (x in c(
  "dplyr", 
  "DT",
  "fs",
  "fst",
  "ggplot2",
  "ggthemes",
  "lubridate",
  "readr",
  "purrr",
  "stringr",
  "tidyr"
  )) {
  suppressPackageStartupMessages(library(x, character.only=TRUE))
}

knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  fig.retina = 2,
  fig.width = 10
)

# TODO: move into the package
source("inc/paths.R")
source("inc/setup.R")
source("inc/latextags.R")

options(repos="https://cran.r-project.org")
theme_set(theme_minimal())
create_tags(path(TAGS_DIR, "side-effects.tex"), prefix="", default=TRUE)
```

```{r}
core_data <- read_fst(EVALS_SUM_CORE_FILE)
package_data <- read_fst(EVALS_SUM_PKGS_FILE)
kaggle_data <- read_fst(EVALS_SUM_KAGGLE_FILE)
program_data <- read_fst(PROGRAM_FILE)

all_data <- rbind(core_data, package_data, kaggle_data)
```

## Side-Effects and Eval

Each entry in `call.fst` contains `direct_writes` and `indirect_writes` (both
`0` for `eval.parent` and `local`). `direct_writes` are the side-effects
directly caused by the `eval` in question and `indirect_writes` are the
side-effects caused by all children evals of the eval. All application
assignments can be found in `direct_writes` field of `program.fst` and all
`eval_assignments` can be found in `indirect_writes` field of `program.fst`. All
non-local assignments can be found by simply summing up the `direct_writes`
field of all eval calls.

```{r preprocess-write-data}
eval_writes_data <-
    all_data %>%
    filter(eval_function == "eval" | eval_function == "evalq") %>%
    select(eval_source_type, eval_source, eval_call_srcref, direct_writes, indirect_writes, nb_ev_calls) %>%
    mutate(all_writes = direct_writes + indirect_writes, side_effecting = as.logical(all_writes))

program_writes_data <-
    program_data %>%
    select(direct_writes, indirect_writes)
```

```{r compute-side-effecting-write-counts}
all_writes <- sum(program_writes_data$direct_writes)
eval_writes <- sum(program_writes_data$indirect_writes)

r("All Writes", all_writes)
r("Eval Writes", eval_writes)

eval_side_effecting_writes <- sum(eval_writes_data$direct_writes * eval_writes_data$nb_ev_calls)

r("Eval Side Effecting Writes", eval_side_effecting_writes)

r("Eval Side Effecting Writes Eval Perc", ratio(eval_side_effecting_writes, eval_writes))
r("Eval Side Effecting Writes All Perc", ratio(eval_side_effecting_writes, all_writes))
```

```{r compute-side-effecting-call-counts}
eval_calls_by_source_writes <-
    eval_writes_data %>%
    count(eval_source_type, side_effecting, wt = nb_ev_calls, name = "call_count") %>%
    group_by(eval_source_type) %>%
    mutate(total_call_count = sum(call_count)) %>%
    ungroup() %>%
    mutate(call_perc = round(100 * call_count/total_call_count, 1)) %>%
    filter(side_effecting)

datatable(eval_calls_by_source_writes)

eval_calls_by_source_writes %>%
pmap(function(eval_source_type, side_effecting, call_count, total_call_count, call_perc) {
    r(str_c("Side Effecting ", eval_source_type, " call perc"), str_c(call_perc, "%"))
})

eval_calls_by_package_writes <-
    eval_writes_data %>%
    filter(eval_source_type != "core") %>%
    count(eval_source, side_effecting, wt = nb_ev_calls, name = "call_count") %>%
    arrange(desc(call_count))

datatable(eval_calls_by_package_writes)

eval_calls_by_source_write_count <-
    eval_writes_data %>%
    filter(all_writes != 0) %>%
    mutate(all_writes = if_else(all_writes <= 5, as.character(all_writes), "> 5")) %>%
    count(eval_source_type, all_writes, wt = nb_ev_calls, name = "call_count") %>%
    group_by(eval_source_type) %>%
    mutate(total_call_count = sum(call_count)) %>%
    ungroup() %>%
    mutate(call_perc = round(100 * call_count/total_call_count, 1))

datatable(eval_calls_by_source_write_count)

plot <-
    eval_calls_by_source_write_count %>%
    ggplot(aes(factor(all_writes, level = c(1:5, "> 5")), call_perc)) +
    geom_col(aes(fill = eval_source_type), position = "dodge") +
    labs(title="", x ="Side-Effects", y = "Side-Effecting Eval Calls %", fill = "") +
    theme(text = element_text(size=15), axis.text.x = element_text(size=15))
plot
```

The call site computation is incorrect.

```{r compute-side-effecting-site-counts, eval = FALSE}
eval_sites_by_source_writes <-
    eval_writes_data %>%
    group_by(eval_source_type, eval_call_srcref) %>%
    summarize(side_effecting = any(side_effecting)) %>%
    ungroup() %>%
    count(eval_source_type, side_effecting, name = "site_count") %>%
    group_by(eval_source_type) %>%
    mutate(total_site_count = sum(site_count)) %>%
    ungroup() %>%
    mutate(site_perc = round(100 * site_count / total_site_count, 1)) %>%
    filter(side_effecting)

datatable(eval_sites_by_source_writes)

eval_sites_by_source_writes %>%
pmap(function(eval_source_type, side_effecting, site_count, total_site_count, site_perc) {
    r(str_c("SideEffecting ", eval_source_type, " site perc"), str_c(site_perc, "%"))
})

eval_sites_by_package_writes <-
    eval_writes_data %>%
    filter(eval_source_type != "core") %>%
    group_by(eval_source, eval_call_srcref) %>%
    summarize(side_effecting = any(side_effecting)) %>%
    ungroup() %>%
    count(eval_source, side_effecting, name = "site_count") %>%
    group_by(eval_source, side_effecting) %>%
    mutate(site_perc = round(100 * site_count / sum(site_count), 1)) %>%
    ungroup() %>%
    arrange(side_effecting, desc(site_perc))

datatable(eval_sites_by_package_writes)
```

>>>>

---
title: "Side-effects"
output:
    html_document:
        gallery: false
        toc: true
        toc_depth: 3
        toc_float: true
        df_print: paged
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
params:
  base_dir: Data
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(DT)
library(digest)
library(fs)
library(fst)
library(purrr)
library(tidyverse)

source("inc/latextags.R")
source("inc/functions.R")
source("inc/paths.R")

create_tags(path(TAGS_DIR, "side-effects.tex"), prefix = "", default = TRUE)
```

```{r load data}
SE <- read_fst(PACKAGE_SIDE_EFFECTS_FILE) %>% as_tibble
```

```{r fix withr::defer callsite}

```

```{r fix S3 methods}

```

```{r}
cs_count(SE, eval_call_srcref, wt=N)
cs_count(SE, variable, wt=N)
```

```{r}
overview_table(
  r("Number of side effects", sum(SE$N))
)
```

## Unknwon srcref

```{r}
unknown_srcref <- filter(SE, eval_call_srcref=="???")
```

## Where does the side effects happen

- in the caller - source is `call:fun` where `fun` is the `calle_function` 
  - most of this should be class that do not specify any parameter to `envir`
- in the supplied `envir` parameter -- `in_envir` is true
  - in this case the details are in the `environment_class`
- in a package - the source is `pkg:`
- in the `.userHookEnv` - the source is `userHookEnv`
- in another call
- is S4, R6

```{r}

```

### 2021-04-12 paper/main.tex

\mypara{Base.} The dataset contains \BasePackages libraries performing basic
arithmetics, statistics, and operating system functionalities. These libraries
are bundled with R and executed pervasively. Together they contain
\BaseEvalCallSites calls to \eval in \BaseFunsWithEvals functions. Some of these
functions provide basic functionalities such as package loading and thus there
is hardly any code which does not invoke one of these.

@@

\begin{wrapfigure}{4}{5.8cm}
\centering
  \begin{tabular}{|r@{\,}r@{\,}l@{\,}r|r@{\,}r@{\,}l@{}r|} \hline
\multicolumn{3}{|c}{\small\#calls} &\small\#sites &
\multicolumn{3}{c}{\small\#calls} &\small\#sites \\\hline
\tt 0 &--& \tt 50    & \packageRunbina & \tt 501 &--& \tt 1000   & \packageRunbine\\
\tt 51 &--& \tt 100  & \packageRunbinb & \tt 1001 &--& \tt 1500  & \packageRunbinf\\
\tt 101 &--& \tt 250 & \packageRunbinc & \tt 1501 &--& \tt 2000  & \packageRunbing\\
\tt 251 &--& \tt 500 & \packageRunbind & \tt 2001 &--& \tt 3000 & \packageRunbinh\\\hline
\end{tabular}

@@

  \includegraphics[width=5.6cm, trim=5.5cm 0 5cm 0, clip]{calls_per_run_per_call_site}

@@

\begin{tabular}{c}
  {\hspace{-3cm}\includegraphics[width=0.8\textwidth]{size_loaded_distribution}}
\end{tabular}
 \caption{Loaded code} \label{fig:sizedistribution}
\end{wrapfigure}

@@

\Eval takes any value, but if its argument is not an expression, \eval returns
it unchanged. Expressions account for \packageCodepercent of arguments in our
corpus. More specifically, \packageSymbolpercent are \c{symbol}s (single
variables such as \c{x}), \packageLanguagepercent are \c{language} object (a
single expression, such as \c{x+1}), and \packageExpressionpercent are
\c{expression} objects (multiple statements). Further inspection reveals that
most symbols, \packageGgplotsymbolpercent to be exact, come from a single call
in the \c{ggplot2} package and have have the value \c{\_inherit}. Even if we
remove this outlier, the majority of \evals are simple expressions.

@@

\subsection{Base}
We recorded \baseAllcalls \eval calls in the base libraries with fewer than 4\%
of those being to any variants (this was done on a small subset of the whole
corpus). Unlike CRAN, Base calls \c{evalq} (\baseEvalqs). Out of \baseStaticeval
call sites, \baseTriggeredevalpct are exercised. Most arguments,
\baseCodepercent, are expressions. However, unlike CRAN, \baseLanguagepercent of
the arguments are language objects. The median argument size is
\baseMedianszeval, which is more than for CRAN. The maximum size is
\baseMaxszeval. Small instruction counts, less than 50, amount for
\baseSmalleventspct of calls. The challenge dealing with Base is that every
program uses it and it can thus generate extreme amounts of data but the data is
quite predictable. For the rest of the paper, we only mention base when there
are surprising observations.

>>>>

\mypara{Base.} The dataset contains \BasePackages libraries performing basic
arithmetics, statistics, and operating system functionalities. These libraries
are bundled with R and executed pervasively. Together they contain
\BaseEvalCallSites call sites to \eval in \BaseFunsWithEvals functions. Some of these
functions provide basic functionalities such as package loading and thus there
is hardly any code which does not invoke one of these.

@@

\begin{wrapfigure}{4}{5.8cm}
  \vspace*{-2mm}
\centering
  \begin{tabular}{|r@{\,}r@{\,}l@{\,}r|r@{\,}r@{\,}l@{}r|} \hline
\multicolumn{3}{|c}{\small\#calls} &\small\#sites &
\multicolumn{3}{c}{\small\#calls} &\small\#sites \\\hline
\tt 0 &--& \tt 50    & \packageRunbina & \tt 501 &--& \tt 1000   & \packageRunbine\\
\tt 51 &--& \tt 100  & \packageRunbinb & \tt 1001 &--& \tt 1500  & \packageRunbinf\\
\tt 101 &--& \tt 250 & \packageRunbinc & \tt 1501 &--& \tt 2000  & \packageRunbing\\
\tt 251 &--& \tt 500 & \packageRunbind & \tt 2001 &--& \tt 3000 & \packageRunbinh\\\hline
\end{tabular}

@@

  \vspace*{-1mm}
  \includegraphics[width=5.6cm, trim=5.5cm 0 5cm 0, clip]{calls_per_run_per_call_site}

@@

\begin{tabular}{c}
  \vspace*{-1mm}
  {\hspace{-3cm}\includegraphics[width=0.8\textwidth]{size_loaded_distribution}}
\end{tabular}
 \caption{Loaded code} \label{fig:sizedistribution}
\end{wrapfigure}

@@

\Eval takes any value, but if its argument is not an expression, \eval returns
it unchanged. Expressions account for \packageCodepercent of arguments in our
corpus. More specifically, \packageSymbolpercent are \texttt{symbol}s (single
variables such as \c{x}), \packageLanguagepercent are \c{language} objects
(function calls such as \c{f(x)} or \c{x+1}), and \packageExpressionpercent are
\c{expression} objects (lists of expressions such as function calls or
symbols). Further inspection reveals that most symbols,
\packageGgplotsymbolpercent to be exact, come from a single site in the
\c{ggplot2} package and have the value \c{\_inherit}.\footnote{This models
inheritance in \c{ggproto}, one of the many object-oriented systems in
R that is used exclusively in the \c{ggplot2} graphics library.}

@@

\subsection{Base}

We recorded \baseAllcalls \eval calls in the \BasePackages base libraries. This
comes from running 10\% randomly selected programs from the
\CranRunnableScriptsRnd extracted programs from CRAN packages, covering
\baseTriggeredevalpct of the \BaseEvalCallSites sites. Most of the calls
(\baseEvalsratio) are to the \c{eval} function and most arguments
(\baseCodepercent) are expressions. However, unlike CRAN, the majority
(\baseLanguagepercent) of the arguments are language objects. The median
argument size is \baseMedianszeval, which is more than for CRAN and the maximum
size is \baseMaxszeval, less than for CRAN. Small instruction counts ($<50$),
amount for \baseSmalleventspct of calls. A single site in the \c{match.arg}
function is responsible for \baseTopFuncPercent of all recorded calls. This
function provides a convenient way for argument verification using partial
matching. It is heavily relied on in R functions that use string arguments to
parameterize function behavior, a popular API mechanism in R. For example, in a
body of \c{center <- function(x, type = c("mean", "median", "trimmed"))} one
can get the value of \c{type} parameter using \c{type <- match.arg(type)}. A
call \c{center(x, "trim")} will match \c{type} to \c{"trimmed"}.

The challenge dealing with Base is that every program uses it and it can thus
generate extreme amounts of data but the data is quite predictable. For the
rest of the paper, we only mention base when there are surprising
observations.
